Variational oblique predictive clustering trees

### 2025-07-17 reports/paper/vspyct_paper.tex
latest changes

In the evolving field of machine learning, the fusion of semi-supervised learning and decision tree models has led to the development of models like \glspl{pct}~\citep{Kocev_2013} and their advanced variant, \glspl{spyct}~\citep{Stepi_nik_2021}.
These models leverage labeled and unlabelled data to enhance predictive accuracy and efficiency.
\Glspl{spyct}, in particular, have pioneered the use of oblique splits for \gls{sop}, allowing for more intricate decision boundaries beyond the capabilities of traditional axis-parallel splits.
\added[id=r1, comment=C4]{In this work we use the term SOP to denote predictive tasks where each example’s label is itself a vector, set or hierarchy (e.g., multi-target regression, hierarchical multi-label, multi-output classification).}
Despite their advantages, a significant challenge with \glspl{spyct} is their reliance on ensemble methods for optimal predictive performance.
Ensemble methods, such as random forests and gradient-boosted trees, achieve high predictive performance by reducing variance and/or bias through the aggregation of diverse models~\cite{Breiman_2001, Kocev_2020, Friedman_2001}.
These methods ensure robustness by capturing slightly different patterns or residuals from the data.
While effective in improving accuracy, ensembles tend to obscure the model's interpretability—a fundamental attribute of decision trees—and do not inherently provide a means to quantify prediction uncertainty, which is crucial for informed decision-making across various applications.

@@

\added[id=r2, comment=C4]{Despite the progress in predictive clustering methods, existing approaches either sacrifice interpretability by relying on ensembles or fail to directly incorporate uncertainty quantification.
Recent works, such as \glspl{spyct}~\citep{Stepi_nik_2021} and \glspl{opct}~\citep{Stepisnik_2020}, have addressed some limitations but lack the integrated capability of uncertainty quantification, critical for real-world decision-making contexts.
Hence, there is a clear research gap in developing methods that simultaneously ensure interpretability, accuracy, and robust uncertainty estimation.}

@@

\added[id=r2, comment=C1C6]{In contrast to prior work, \glspl{vspyct} combine oblique splits from predictive clustering trees with a Bayesian inference framework using variational Bayes.
Unlike traditional ensemble methods or single deterministic trees, \glspl{vspyct} explicitly model uncertainty in split parameters, providing inherent uncertainty quantification without sacrificing interpretability.
This integration represents a significant novelty by directly embedding probabilistic modeling into oblique splits, ensuring a single interpretable model can achieve performance comparable to or surpassing state-of-the-art ensembles across various prediction tasks.}

@@

\added[id=r1, comment=C8]{The complete implementation is available at \url{https://github.com/viksa98/variational_spyct}.}

@@

\added[id=r2, comment=C14]{Advanced modeling methods, including fractional calculus and artificial neural networks, have been applied extensively across diverse domains for solving complex modeling and prediction problems~\cite{Abbas_2024,Abbas_2025,Ramzan_2025,Abbas_2025_2, Andonovikj_2022}.}

@@

\added[id=r1, comment=C3]{To provide a clearer overview of the related predictive clustering approaches discussed, we summarise and contrast their interpretability, capability to handle high-dimensional data, and uncertainty quantification properties in Table~\ref{tab:model_comparison}.}

@@

\added[id=r2, comment=C8C9]{We selected Gaussian priors for split parameters due to their computational tractability and effectiveness in Bayesian inference frameworks.
The impurity function used for evaluating splits leverages the variance reduction weighted by selection probabilities, a common choice ensuring effective partitioning of data into homogeneous subsets.
Feature importance is calculated using the normalised expected weight-to-variance ratio, reflecting both the strength and reliability of each feature's contribution.}

@@

\subsection{Parameter Configuration Summary}
\added[id=r2, comment=C12]{}

@@

\added[id=r1, comment=C7]{While the present work marginalises only over split-parameter posteriors, future work will adopt Bayesian non-parametric priors over tree topologies to capture structural, as well as parameter, uncertainty. This would allow the model to reason not only about decision-boundary uncertainty, but also about the tree structure itself (i.e., how many splits to create and where to place them). Possible approaches include integrating structural priors, or developing ensemble-like variational methods that maintain a distribution over multiple plausible tree structures. Such extensions would provide a more comprehensive Bayesian treatment of decision trees and further enhance both predictive reliability and interpretability.}

\added[id=r1, comment=C6]{An explicit empirical comparison of training time and scalability between \gls{vspyct} and deterministic approaches was beyond the scope of this study, due to implementation-specific and hardware-dependent factors.}

\textit{Advantages of VSPYCT:}
\added[id=r2, comment=C10]{}
\begin{itemize}
    \item Integrated interpretability and uncertainty quantification.
    \item Competitive predictive performance compared to ensembles.
    \item Effective handling of complex classification boundaries.
\end{itemize}

@@

\begin{figure}[h!]
    \centering
    \includegraphics{../figures/feat_importances.pdf}
    \caption{Feature importance scores for both the original and noise-augmented features.
    Blue markers represent the importance of original features, while orange markers correspond to noise features. The horizontal threshold line marks the point where removing additional features starts degrading model performance. Features below this threshold are non-essential, as their removal improves the evaluation score—demonstrating that noise features (orange) contribute little to model performance. In contrast, features above the threshold are crucial, as removing them further degrades performance.}
    \label{fig:feature_importances}
\end{figure}

@@

\subsection{Practical Interpretation of Results}
\added[id=r2, comment=C7]{}
The empirical results presented earlier can be further understood by considering their practical implications and physical meaning in real-world contexts.
For regression tasks, such as predicting concrete compressive strength, the improved accuracy of \gls{vspyct} indicates that probabilistic modeling of uncertainty in feature weights enables better handling of measurement variability or experimental errors common in physical systems.
This modeling approach allows practitioners, for example civil engineers, to better anticipate the reliability of material properties predicted by the model.

@@

\added[id=r1, comment=C9]{}
After training, VSPYCT provides three complementary outputs that can be inspected with standard analysis tools:
\begin{enumerate}
  \item \textbf{Node summaries:} for every split we report the posterior mean, standard deviation, and $95\%$ credible interval of each coefficient, allowing analysts to verify which variables dominate within a specific branch.
  \item \textbf{Tree file:} the complete tree is exported in a plain, machine-readable graph format so that users can open it in any network-visualisation environment and interactively traverse, fold, or unfold decision paths.
  \item \textbf{Global relevance ranking:} the importance scores (see Figure~\ref{fig:feature_importance}) order the features by their expected effect on the prediction, helping experts prioritise factors for monitoring or intervention.
\end{enumerate}

@@

\added[id=r2, comment=C2C13]{}
We introduced \glspl{vspyct}, a novel framework combining Bayesian inference with predictive clustering trees.
\Glspl{vspyct} uniquely incorporate variational Bayes into oblique splits, providing built-in uncertainty quantification and preserving interpretability.

>>>>

In the evolving field of machine learning, the fusion of semi-supervised learning and decision tree models has led to the development of models like \glspl{pct}~\citep{Kocev_2013} and their advanced variant, \glspl{spyct}~\citep{Stepi_nik_2021}.
These models leverage labeled and unlabelled data to enhance predictive accuracy and efficiency.
\Glspl{spyct}, in particular, have pioneered the use of oblique splits for \gls{sop}, allowing for more intricate decision boundaries beyond the capabilities of traditional axis-parallel splits.
In this work we use the term SOP to denote predictive tasks where each example’s label is itself a vector, set or hierarchy (e.g., multi-target regression, hierarchical multi-label, multi-output classification).
Despite their advantages, a significant challenge with \glspl{spyct} is their reliance on ensemble methods for optimal predictive performance.
Ensemble methods, such as random forests and gradient-boosted trees, achieve high predictive performance by reducing variance and/or bias through the aggregation of diverse models~\cite{Breiman_2001, Kocev_2020, Friedman_2001}.
These methods ensure robustness by capturing slightly different patterns or residuals from the data.
While effective in improving accuracy, ensembles tend to obscure the model's interpretability—a fundamental attribute of decision trees—and do not inherently provide a means to quantify prediction uncertainty, which is crucial for informed decision-making across various applications.

@@

Despite the progress in predictive clustering methods, existing approaches either sacrifice interpretability by relying on ensembles or fail to directly incorporate uncertainty quantification.
Recent works, such as \glspl{spyct}~\citep{Stepi_nik_2021} and \glspl{opct}~\citep{Stepisnik_2020}, have addressed some limitations but lack the integrated capability of uncertainty quantification, critical for real-world decision-making contexts.
Hence, there is a clear research gap in developing methods that simultaneously ensure interpretability, accuracy, and robust uncertainty estimation.

@@

In contrast to prior work, \glspl{vspyct} combine oblique splits from predictive clustering trees with a Bayesian inference framework using variational Bayes.
Unlike traditional ensemble methods or single deterministic trees, \glspl{vspyct} explicitly model uncertainty in split parameters, providing inherent uncertainty quantification without sacrificing interpretability.
This integration represents a significant novelty by directly embedding probabilistic modeling into oblique splits, ensuring a single interpretable model can achieve performance comparable to or surpassing state-of-the-art ensembles across various prediction tasks.

@@

The complete implementation is available at \url{https://github.com/viksa98/variational_spyct}.

@@

Advanced modeling methods, including fractional calculus and artificial neural networks, have been applied extensively across diverse domains for solving complex modeling and prediction problems~\cite{Abbas_2024,Abbas_2025,Ramzan_2025,Abbas_2025_2, Andonovikj_2022}.

@@

To provide a clearer overview of the related predictive clustering approaches discussed, we summarise and contrast their interpretability, capability to handle high-dimensional data, and uncertainty quantification properties in Table~\ref{tab:model_comparison}.

@@

We selected Gaussian priors for split parameters due to their computational tractability and effectiveness in Bayesian inference frameworks.
The impurity function used for evaluating splits leverages the variance reduction weighted by selection probabilities, a common choice ensuring effective partitioning of data into homogeneous subsets.
Feature importance is calculated using the normalised expected weight-to-variance ratio, reflecting both the strength and reliability of each feature's contribution.

@@

\subsection{Parameter Configuration Summary}

@@

While the present work marginalises only over split-parameter posteriors, future work will adopt Bayesian non-parametric priors over tree topologies to capture structural, as well as parameter, uncertainty. This would allow the model to reason not only about decision-boundary uncertainty, but also about the tree structure itself (i.e., how many splits to create and where to place them). Possible approaches include integrating structural priors, or developing ensemble-like variational methods that maintain a distribution over multiple plausible tree structures. Such extensions would provide a more comprehensive Bayesian treatment of decision trees and further enhance both predictive reliability and interpretability.

An explicit empirical comparison of training time and scalability between \gls{vspyct} and deterministic approaches was beyond the scope of this study, due to implementation-specific and hardware-dependent factors.

\textit{Advantages of VSPYCT:}
\begin{itemize}
    \item Integrated interpretability and uncertainty quantification.
    \item Competitive predictive performance compared to ensembles.
    \item Effective handling of complex classification boundaries.
\end{itemize}

@@

\begin{figure}[h!]
    \centering
    \includegraphics{feat_importances.pdf}
    \caption{Feature importance scores for both the original and noise-augmented features.
    Blue markers represent the importance of original features, while orange markers correspond to noise features. The horizontal threshold line marks the point where removing additional features starts degrading model performance. Features below this threshold are non-essential, as their removal improves the evaluation score—demonstrating that noise features (orange) contribute little to model performance. In contrast, features above the threshold are crucial, as removing them further degrades performance.}
    \label{fig:feature_importances}
\end{figure}

@@

\subsection{Practical Interpretation of Results}
The empirical results presented earlier can be further understood by considering their practical implications and physical meaning in real-world contexts.
For regression tasks, such as predicting concrete compressive strength, the improved accuracy of \gls{vspyct} indicates that probabilistic modeling of uncertainty in feature weights enables better handling of measurement variability or experimental errors common in physical systems.
This modeling approach allows practitioners, for example civil engineers, to better anticipate the reliability of material properties predicted by the model.

@@

After training, VSPYCT provides three complementary outputs that can be inspected with standard analysis tools:
\begin{enumerate}
  \item \textbf{Node summaries:} for every split we report the posterior mean, standard deviation, and $95\%$ credible interval of each coefficient, allowing analysts to verify which variables dominate within a specific branch.
  \item \textbf{Tree file:} the complete tree is exported in a plain, machine-readable graph format so that users can open it in any network-visualisation environment and interactively traverse, fold, or unfold decision paths.
  \item \textbf{Global relevance ranking:} the importance scores (see Figure~\ref{fig:feature_importance}) order the features by their expected effect on the prediction, helping experts prioritise factors for monitoring or intervention.
\end{enumerate}

@@

We introduced \glspl{vspyct}, a novel framework combining Bayesian inference with predictive clustering trees.
\Glspl{vspyct} uniquely incorporate variational Bayes into oblique splits, providing built-in uncertainty quantification and preserving interpretability.
